
# Coursera: Financial Engineering I: In-Class Examples in R

Course page: [https://www.coursera.org/learn/financial-engineering-1](https://www.coursera.org/learn/financial-engineering-1)

* [Lesson 3.1: Mechanics of the Futures Margin Account](#3.1)
* [Lesson 3.2: Stock Price Dynamics in the Binomial Model](#3.2)
* [Lesson 3.4: Option pricing in the 1-period binomial model](#3.4)
* [Lesson 4.1: Option pricing in the multi-period binomial model](#4.1)
* [Lesson 4.2: Pricing American Options](#4.2)
* [Lesson 4.4: Pricing a European Put on a Futures Contract](#4.4)
* [Lesson 5.1: Pricing a ZCB ](#5.1)
* [Lesson 5.2: Fixed-Income Derivatives](#5.2)
* [Lesson 5.3: Caplets, Floorlets, Swaps, Swaptions](#5.3)
* [Lesson 5.4: The Forward Equations (Elementary Securities)](#5.4)
* [Lesson 6.1: Black-Derman-Toy (BDT) Model Calibration](#6.1)
* [Lesson 6.2: Pricing using Calibrated BDT Model](#6.2)
* [Lesson 6.3: Pricing Defaultable Bonds, Calibrating Hazard Rates](#6.3)
* [Lesson 6.5: Pricing Credit Default Swaps](#6.5)
* [Lesson 7.1: Mortgage mathematics](#7.1)
* [Lesson 7.2: MBS Pass Thru Mathematics](#7.2)
* [Lesson 7.4: MBS Pass-Thru Sequential CMO with Tranches](#7.4)

```{r}

    source("fin-eng-1.R")

```

<br /><br />
## <a name="3.1"></a>Lesson 3.1: Mechanics of the Futures Margin Account

### Corn Futures Margin Account

```{r}

    spot.prices <- c(690.25,
                     693.50,
                     705.00,
                     709.50,
                     719.50,
                     724.25,
                     723.00,
                     732.00,
                     708.00,
                     711.50,
                     725.25,
                     734.50,
                     741.00,
                     741.25,
                     732.75,
                     730.50) / 100

    margin.schedule <- buildMarginAccountSchedule( spot.price.vec=spot.prices,
                                                   num.contracts=5000,
                                                   initial.margin=1688,
                                                   maintenance.margin=1250 )

    data.frame(margin.schedule)

    sum(margin.schedule$profit)


```


### Random Corn Futures Margin Account w/ Margin Call

```{r}

    # (random) spot prices, quoted in dollars per bushel
    spot.prices <- c(690.25,
                     686.81,
                     686.59,
                     668.51,
                     662.08,
                     674.89,
                     661.97,
                     709.88,
                     719.20,
                     709.36,
                     725.84,
                     743.18,
                     733.77,
                     743.12,
                     776.84,
                     796.17) / 100

    margin.schedule <- buildMarginAccountSchedule( spot.price.vec=spot.prices,
                                                   num.contracts=5000,
                                                   initial.margin=1688,
                                                   maintenance.margin=1250 )

    data.frame(margin.schedule)

    sum(margin.schedule$profit)

```



<br /><br />
## <a name="3.2"></a>Lesson 3.2: Stock Price Dynamics in the Binomial Model

```{r}


    stock.price.lattice <- buildAssetPriceLattice(S0=100,
                                                  n=3,
                                                  u=1.07,
                                                  d=1/1.07) 
    stock.price.lattice

```

<br /><br />
## <a name="3.4"></a>Lesson 3.4: Option pricing in the 1-period binomial model

```{r}
 
    u = 1.07
    d = 1/u
    R = 1.01
    q = (R - d) / (u - d)

    option.price.lattice <- buildOptionPriceLattice(asset.price.lattice=stock.price.lattice,
                                                    n=1,
                                                    K=102,
                                                    R=1.01,
                                                    q=q,
                                                    is.call.option=T,
                                                    is.amer.option=F)
    option.price.lattice 


```


<br /><br />
## <a name="4.1"></a>Lesson 4.1: Option pricing in the multi-period binomial model

```{r}

    option.price.lattice <- buildOptionPriceLattice(asset.price.lattice=stock.price.lattice,
                                                    n=3,
                                                    K=100,
                                                    R=1.01,
                                                    q=q,
                                                    is.call.option=T,
                                                    is.amer.option=F)
    option.price.lattice 

```

<br /><br />
## <a name="4.2"></a>Lesson 4.2: Pricing American Options



```{r}

    option.price.lattice <- buildOptionPriceLattice(asset.price.lattice=stock.price.lattice,
                                                    n=3,
                                                    K=100,
                                                    R=1.01,
                                                    q=q,
                                                    is.call.option=F,
                                                    is.amer.option=T)
    option.price.lattice 

    shouldExerciseOption( asset.price.lattice = stock.price.lattice,
                          option.price.lattice = option.price.lattice,
                          K=100,
                          is.call.option = F )

```


<br /><br />
## <a name="4.4"></a>Lesson 4.4: Pricing a European Put on a Futures Contract

#### Step 1: Calibrate a Binomial Model via Black-Scholes Parameters

```{r}

    T = 0.50
    sigma = 20/100
    n = 10
    r = 2/100
    c = 1/100

    bin.model <- calibrateBinomialModel(T=T,
                                        n=n,
                                        r=r,
                                        c=c,
                                        sigma=sigma)

    rbind(bin.model)


```

#### Step 2: Build stock price lattice

```{r}


    stock.price.lattice <- buildAssetPriceLattice(S0=100,
                                                  n=n,
                                                  u=bin.model$u,
                                                  d=bin.model$d) 
    stock.price.lattice

```


#### Step 3: Build Futures price lattice

```{r}

    futures.price.lattice <- buildFuturesPriceLattice(asset.price.lattice=stock.price.lattice,
                                                      n=n,
                                                      q=bin.model$q)

    futures.price.lattice

```

#### Step 4: Build Put Option price lattice

```{r}

    option.price.lattice <- buildOptionPriceLattice(asset.price.lattice=futures.price.lattice,
                                                    n=n,
                                                    K=100,
                                                    R=bin.model$R,
                                                    q=bin.model$q,
                                                    is.call.option=F,
                                                    is.amer.option=F)
    option.price.lattice

```




<br /><br />
## <a name="5.1"></a>Lesson 5.1: Pricing a ZCB 


#### Step 1: Build A Sample Short Rate Lattice

```{r}


    short.rate.lattice <- buildShortRateLattice(r0=0.06,
                                                n=5,
                                                u=1.25,
                                                d=0.9)

    short.rate.lattice

```

#### Step 2: Price ZCB using short-rate lattice

```{r}

    zcb.price.lattice <- buildZCBPriceLattice(short.rate.lattice = short.rate.lattice,
                                              face.value = 100,
                                              n=4,
                                              q=0.5)
    zcb.price.lattice

```


<br /><br />
## <a name="5.2"></a>Lesson 5.2: Fixed-Income Derivatives


### Pricing a Euro Call Option on ZCB:

```{r}

    option.price.lattice <- buildOptionPriceLattice(asset.price.lattice=zcb.price.lattice,
                                                    n=2,
                                                    K=84,
                                                    short.rate.lattice = short.rate.lattice,
                                                    q=0.5,
                                                    is.call.option=T,
                                                    is.amer.option=F)
    option.price.lattice 

```

### Pricing an American Put Option on ZCB:

```{r}

    option.price.lattice <- buildOptionPriceLattice(asset.price.lattice=zcb.price.lattice,
                                                    n=3,
                                                    K=88,
                                                    short.rate.lattice = short.rate.lattice,
                                                    q=0.5,
                                                    is.call.option=F,
                                                    is.amer.option=T)
    option.price.lattice 

    shouldExerciseOption( asset.price.lattice = zcb.price.lattice,
                          option.price.lattice = option.price.lattice,
                          K=88, 
                          is.call.option = F )

```


### Pricing a Forward on a Coupon-Bearing Bond:



#### Step 1: Price the Coupon-Bearing Bond:

```{r}

    bond.price.lattice <- buildBondPriceLattice( short.rate.lattice = short.rate.lattice,
                                                 face.value = 100,
                                                 coupon.rate = 0.10,
                                                 n = 6,
                                                 q = 0.5)
    bond.price.lattice

    # ex-coupon at t=4:
    ex.coupon.price <- bond.price.lattice[,4+1] - 100 * 0.10
    t(t(ex.coupon.price))

```

#### Step 2: Build Forward Price Lattice:

```{r}

    forward.price.lattice <- buildBinomialPriceLattice( final.price.vec = ex.coupon.price,
                                                        short.rate.lattice = short.rate.lattice,
                                                        n = 4,
                                                        q = 0.5)
    forward.price.lattice

```

#### Step 3: Price ZCB with T=4 and face value=1:

```{r}

    zcb.price.lattice <- buildZCBPriceLattice(short.rate.lattice = short.rate.lattice,
                                              face.value = 1,
                                              n=4,
                                              q=0.5)
    zcb.price.lattice

```

#### Step 4: Compute fair-value of forward price

```{r}

    forward.price.lattice[1,1] / zcb.price.lattice[1,1]

```


### Pricing a Futures on a Coupon-Bearing Bond:


```{r}

    futures.price.lattice <- buildFuturesPriceLattice( final.price.vec = ex.coupon.price,
                                                       n = 4,
                                                       q = 0.5)
    futures.price.lattice

```



<br /><br />
## <a name="5.3"></a>Lesson 5.3: Caplets, Floorlets, Swaps, Swaptions


### Pricing a Caplet

```{r}

    K <- 0.02

    # final cash flow is discounted (paid in arrears)
    caplet.final.price <- pmax( (short.rate.lattice[,6] - K) / (1 + short.rate.lattice[,6]), 0 )

    caplet.price.lattice <- buildBinomialPriceLattice(final.price.vec = caplet.final.price,
                                                      short.rate.lattice = short.rate.lattice,
                                                      n = 5,  # technically expires at t=6
                                                      q=0.5)
    caplet.price.lattice 

```

### Pricing an Interest Rate Swap

```{r}

    swap.price.lattice <- buildInterestRateSwapPriceLattice( short.rate.lattice = short.rate.lattice,
                                                             n = 5,  # technically expires at t=6
                                                             q = 0.5,
                                                             K = 0.05) 
    swap.price.lattice
    
```

### Pricing a Swaption

```{r}

    swaption.price.lattice <- buildOptionPriceLattice(asset.price.lattice = swap.price.lattice,
                                                      n=3,
                                                      K=0,
                                                      short.rate.lattice = short.rate.lattice,
                                                      q=0.5,
                                                      is.call.option=T,
                                                      is.amer.option=F)
    swaption.price.lattice

```


<br /><br />
## <a name="5.4"></a>Lesson 5.4: The Forward Equations (Elementary Securities)


### Building the Elementary Security Price Lattice

```{r}


    elem.price.lattice <- buildElementarySecurityPriceLattice(short.rate.lattice = short.rate.lattice,
                                                              n = 6,
                                                              q=0.5)
    elem.price.lattice

```

### Pricing a ZCB using Elementary Prices

```{r}

    # all ZCB bond prices, w/ expirations t=0 thru t=6
    zcb.prices <- colSums( elem.price.lattice, na.rm=T) * 100
    zcb.prices

    # spot rates
    (100/zcb.prices)^(1/0:6) - 1

```

### Pricing a Forward-Start Swap using Elementary Prices

```{r}

    swap.coupon.lattice <- buildInterestRateSwapCouponPaymentLattice( short.rate.lattice = short.rate.lattice,
                                                                      n = 2,    # technically expires at t=3
                                                                      forward.start = 1,
                                                                      K = 0.07) 
    swap.coupon.lattice

    discounted.swap.coupon.lattice <- swap.coupon.lattice / (1 + short.rate.lattice[1:3,1:3])
    discounted.swap.coupon.lattice 

    sum( discounted.swap.coupon.lattice * elem.price.lattice[1:3,1:3], na.rm=T ) * 1e6

```


<br /><br />
## <a name="6.1"></a>Lesson 6.1: Black-Derman-Toy (BDT) Model Calibration

```{r}

    q <- 0.5
    n <- 13
    s <- c(7.3, 7.62, 8.1, 8.45, 9.2, 9.64, 10.12, 10.45, 10.75, 11.22, 11.55, 11.92, 12.2, 12.32) / 100
    a <- rep(5.0/100,n+1)       # starting values for a (to be calibrated)
    b <- rep(0.5/100,n+1)       # fixed b for all time periods

    o <- calibrateBDTShortRateLattice(a=a,
                                      b=b,
                                      s=s,
                                      q=q,
                                      n=n,     # num periods in calibrated BDT short-rate lattice
                                      optim.rounds=1)
    o

```

<br /><br />
## <a name="6.2"></a>Lesson 6.2: Pricing using Calibrated BDT Model

### Pricing a forward-start swaption

```{r}

    forward.start.swap.lattice <- buildInterestRateSwapPriceLattice( short.rate.lattice=o$bdt,
                                                                     n=9,   # technically expires at t=10
                                                                     forward.start=2,
                                                                     K=11.65/100 )
    forward.start.swap.lattice 

    swaption.price.lattice <- buildOptionPriceLattice( forward.start.swap.lattice,
                                                       K=0,
                                                       n=2,
                                                       short.rate.lattice=o$bdt )
    swaption.price.lattice
    swaption.price.lattice[1,1] * 1e6

```
**Note**: answer is slightly off from lecture notes, due to diffs in calibration

### Now change b (volatility) to 0.01:

```{r}

    b <- rep(1.0/100,n+1)

    o <- calibrateBDTShortRateLattice(a=a,
                                      b=b,
                                      s=s,
                                      q=q,
                                      n=n,
                                      optim.rounds=1)
    o


    forward.start.swap.lattice <- buildInterestRateSwapPriceLattice( short.rate.lattice=o$bdt,
                                                                     n=9,   # technically expires at t=10
                                                                     forward.start=2,
                                                                     K=11.65/100 )
    forward.start.swap.lattice 

    swaption.price.lattice <- buildOptionPriceLattice( forward.start.swap.lattice,
                                                       K=0,
                                                       n=2,
                                                       short.rate.lattice=o$bdt )
    swaption.price.lattice
    swaption.price.lattice[1,1] * 1e6

```

**Note**: answer is slightly off from lecture notes, due to diffs in calibration


<br /><br />
## <a name="6.3"></a>Lesson 6.3: Pricing Defaultable Bonds, Calibrating Hazard Rates


### Pricing a Defaultable Bond


```{r}

    delta = 1/2             # coupon payments every 6 mos.
    h <- rep(0.02,10+1)     # fixed hazard rate

    bond.1yr <- buildDefaultableBondSchedule( face.value = 100,
                                              hazard.rate.vec = h,
                                              r = 0.05,
                                              recovery.rate = 0.10,
                                              coupon.rate = 0.05,
                                              delta = delta,
                                              n = 1/delta )
    
    data.frame(bond.1yr)
    
    bond.1yr.price <- sum(bond.1yr$all.payment)
    bond.1yr.price


    bond.5yr <- buildDefaultableBondSchedule( face.value = 100,
                                              hazard.rate.vec = h,
                                              r = 0.05,
                                              recovery.rate = 0.20,
                                              coupon.rate = 0.10,
                                              delta = delta,
                                              n = 5 / delta )
    data.frame(bond.5yr)
    
    bond.5yr.price <- sum(bond.5yr$all.payment)
    bond.5yr.price

```


### Calibrating Hazard Rates

```{r}

    # these are just random values
    market.prices <- c(101.2179, 
                       92.5828, 
                       107.3675, 
                       104.0419, 
                       145.9155)


    #
    # compute 1-5 yr bond prices using the given hazard rates
    # 
    computeBondPrices <- function(h) {

        delta = 1/2     # coupon payments every 6 mos.

        bond.1yr <- computeDefaultableBondPricing( face.value = 100,
                                                   hazard.rate.vec = h,
                                                   r = 0.05,
                                                   recovery.rate = 0.10,
                                                   coupon.rate = 0.05,
                                                   delta = delta,
                                                   n = 1/delta )

        bond.2yr <- computeDefaultableBondPricing( face.value = 100,
                                                   hazard.rate.vec = h,
                                                   r = 0.05,
                                                   recovery.rate = 0.25,
                                                   coupon.rate = 0.02,
                                                   delta = delta,
                                                   n = 2 / delta )

        bond.3yr <- computeDefaultableBondPricing( face.value = 100,
                                                   hazard.rate.vec = h,
                                                   r = 0.05,
                                                   recovery.rate = 0.50,
                                                   coupon.rate = 0.05,
                                                   delta = delta,
                                                   n = 3 / delta )

        bond.4yr <- computeDefaultableBondPricing( face.value = 100,
                                                   hazard.rate.vec = h,
                                                   r = 0.05,
                                                   recovery.rate = 0.10,
                                                   coupon.rate = 0.05,
                                                   delta = delta,
                                                   n = 4 / delta )

        bond.5yr <- computeDefaultableBondPricing( face.value = 100,
                                                   hazard.rate.vec = h,
                                                   r = 0.05,
                                                   recovery.rate = 0.20,
                                                   coupon.rate = 0.10,
                                                   delta = delta,
                                                   n = 5 / delta )

        return( list(bond.1yr=bond.1yr, 
                     bond.2yr=bond.2yr, 
                     bond.3yr=bond.3yr, 
                     bond.4yr=bond.4yr, 
                     bond.5yr=bond.5yr) )
    }

    # 
    # define objective function (to be minimized)
    # 
    objectiveFn <- function(h) {

        bonds <- computeBondPrices(h)

        model.prices <- c(bonds$bond.1yr$price, 
                          bonds$bond.2yr$price, 
                          bonds$bond.3yr$price, 
                          bonds$bond.4yr$price, 
                          bonds$bond.5yr$price )

        sq.diffs <- (market.prices - model.prices)^2

        sum(sq.diffs)       # to minimize
    }



    o <- optim(par=h,
               fn=objectiveFn,
               method = "L-BFGS-B")
    o

    bonds <- computeBondPrices(o$par)

    rbind(model.prices=c(bonds$bond.1yr$price, bonds$bond.2yr$price, bonds$bond.3yr$price, bonds$bond.4yr$price, bonds$bond.5yr$price), 
          market.prices)

```

### Calibration using R constrOptim():

```{r}

    # 
    # using a constraint: allow positive values only.
    # 
    # ui %*% theta - ci >= 0
    # kxp     px1    kx1
    # 
    ui <- matrix(1,nrow=1,ncol=length(h))
    ci <- c(0)
    ui %*% h - ci   

    o1 <- constrOptim( theta=h,
                       f=objectiveFn,
                       grad=NULL,
                       ui=ui,
                       ci=ci)
    o1

    bonds <- computeBondPrices(o1$par)

    rbind(model.prices=c(bonds$bond.1yr$price, bonds$bond.2yr$price, bonds$bond.3yr$price, bonds$bond.4yr$price, bonds$bond.5yr$price), 
          market.prices)

```

### Calibration using R constrOptim(): Non-decreasing hazard rates

```{r}

    # 
    # using a constraint: pair-wise compare hazard rates
    # to ensure non-decreasing
    # 
    # ui %*% theta - ci >= 0
    # kxp     px1    kx1
    # 
    ui <- matrix(c(
                   -1,  1,  0,  0,  0,  0,  0,  0,  0,  0,  0,
                    0, -1,  1,  0,  0,  0,  0,  0,  0,  0,  0,
                    0,  0, -1,  1,  0,  0,  0,  0,  0,  0,  0,
                    0,  0,  0, -1,  1,  0,  0,  0,  0,  0,  0,
                    0,  0,  0,  0, -1,  1,  0,  0,  0,  0,  0,
                    0,  0,  0,  0,  0, -1,  1,  0,  0,  0,  0,
                    0,  0,  0,  0,  0,  0, -1,  1,  0,  0,  0,
                    0,  0,  0,  0,  0,  0,  0, -1,  1,  0,  0,
                    0,  0,  0,  0,  0,  0,  0,  0, -1,  1,  0,
                    0,  0,  0,  0,  0,  0,  0,  0,  0, -1,  1 ),
                 nrow = 10,
                 ncol = 11,
                 byrow=T)
    ci <- rep(0,10)
    h1 <- h + cumsum(rep(0.0001,length(h))) - 0.0005
    ui %*% h1 - ci   

    o2 <- constrOptim( theta=h1,
                       f=objectiveFn,
                       grad=NULL,
                       ui=ui,
                       ci=ci)
    o2

    bonds <- computeBondPrices(o2$par)

    rbind(model.prices=c(bonds$bond.1yr$price, bonds$bond.2yr$price, bonds$bond.3yr$price, bonds$bond.4yr$price, bonds$bond.5yr$price), 
          market.prices)

```

<br /><br />
## <a name="6.5"></a>Lesson 6.5: Pricing Credit Default Swaps

```{r}

    # copied from spreadsheet
    h <- c(0.010070286520099,
           0.010070286520099,
           0.010070286520099,
           0.010070286520099,
           0.009721551669331,
           0.009721551669331,
           0.009721551669331,
           0.009721551669331,
           0.009721551669331)      # for time periods t=0 thru n


    cds <- buildCDSSchedule( N = 1e6,                 # principal
                             hazard.rate.vec = h,
                             n = 8,
                             delta = 1/4,
                             r = 0.01,
                             recovery.rate = 0.45 )
    data.frame(cds)

```



<br /><br />
## <a name="7.1"></a>Lesson 7.1: Mortgage mathematics


```{r}

    mort <- buildMortgageSchedule(mortgage.balance = 20000,
                                  monthly.mortgage.rate = 0.05/12,
                                  n = 18)

    data.frame(mort)

    computeMortgagePresentValue(monthly.payment = mort$monthly.payment[1],
                                monthly.market.rate = 0.05/12,
                                n = 18)

```

<br /><br />
## <a name="7.2"></a>Lesson 7.2: MBS Pass-Thru Mathematics

```{r}

    cpr.rate <- buildCPRSchedule( psa.multiplier = 100,
                                  seasoning = 0,
                                  n = 30 )

    mbs <- buildMBSPassThruSchedule( mortgage.balance = 400,
                                     monthly.mortgage.rate = 8.125/100/12,
                                     monthly.passthru.rate = 7.500/100/12,
                                     cpr.rate = cpr.rate[4:20],
                                     n = 17) 
    data.frame(mbs)

    avg.life <- 1 / 12 / 400 * sum( 1:17 * mbs$total.principal )
    avg.life
    
```

<br /><br />
## <a name="7.4"></a>Lesson 7.4: MBS Pass-Thru Sequential CMO with Tranches 


```{r}

    tranches <- buildMBSTrancheSchedule( mbs=mbs,
                                         tranche.begin.balances = c(194.5, 36, 96.5, 73),
                                         n=17 )

    data.frame(tranches[1])
    data.frame(tranches[2])
    data.frame(tranches[3])
    data.frame(tranches[4])

```
